An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-02-07 implementation.tex
Updates from Overleaf

\vspace{1ex}\noindent {\bf Example properties.}
Figure \ref{fig:properties} shows three properties in the 
input format of the tool, which are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. The properties are related to the examples in Section~\ref{sec:syntax-semantics}, which
are not expressible in (P)FLTL.
%
The ASCII version of the logic here uses 
\idsl{@} for $\ominus$,
\idsl{|} for $\vee$,
\idsl{&} for $\wedge$, and
\idsl{!} for $\neg$.
%
The first property \idsl{telemetry1} is a variant of
formula \ref{eq:wolper-first-order}, illustrating the use of a 
rule to express a first-order version of Wolper's example \cite{Wolper}, that all the states with even indexes of a sequence satisfy a property.
In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula), which can be turned on and off with a \idsl{toggle(x)} - they are initially off.
Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when radio channel \idsl{x} is toggled on.

>>>>

\vspace{1ex}\noindent {\bf Example properties.}
Figure \ref{fig:properties} shows three properties in the 
input format of the tool, related to the examples in Section~\ref{sec:syntax-semantics}, which
are not expressible in (P)FLTL.
These properties are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. The properties are 
%
The ASCII version of the logic here uses 
\idsl{@} for $\ominus$,
\idsl{|} for $\vee$,
\idsl{&} for $\wedge$, and
\idsl{!} for $\neg$.
%
The first property \idsl{telemetry1} is a variant of
formula \ref{eq:wolper-first-order}, illustrating the use of a 
rule to express a first-order version of Wolper's example \cite{Wolper}, that all the states with even indexes of a sequence satisfy a property.
In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula), which can be turned on and off with a \idsl{toggle(x)} - they are initially off.
Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when radio channel \idsl{x} is toggled on.

### 2019-02-07 rule-ltl.tex
Updates from Overleaf

\item Observe a new event (as a set propositions) $s$ as input. 
\item Let $\old := \current$.
\item Make the following updates for each subformula. If $\varphi$ is
      a subformula of $\psi$ then $\current ( \varphi )$ is updated before 
      $\current ( \psi )$.
\begin{itemize}
  \item $\current ( \true ) := \true$.
  \item $\current (  \varphi \wedge \psi  ) := 
  \current ( \varphi )\  and\ \current ( \psi )$.
  \item $\current ( \neg \varphi  ) := not\ \current ( \varphi )$.
  \item $\current (  \varphi \since \psi  ) :=  
  \current ( \psi  )\ or\ ( \current ( \varphi ) \ and\ 
      \old ( ( \varphi \since  \psi )))$.
%DAP: I removed the \; around \since, since it penetrated the 2nd column!
  \item $\current ( \ominus \; \varphi ) := \old ( \varphi )$.
\end{itemize}
\item If $\current (\eta) = \false$ then
report a violation and terminate.
\item Goto step~2.
\end{enumerate}

@@

\noindent
We define\footnote{ Formal semantics 
can also be given by constructing a set of temporal relations extended
with the auxiliary ones
inductively over growing prefixes, as would be done in a detailed proof of Lemma~\ref{samesame}.} the semantics for the EFLTL 
specification (\ref{EFLTL}) 
by using the following equivalent
QFLTL formula:

@@

\noindent {\bf Sketch of Proof.}
The proof of this theorem, which includes encoding of a nontrivial property, where the property
observes the data elements (appearing as $v (a)$)
in between states where $r$ appears (a relation with
one parameter, the time or index parameter $i$).
The property asserts that there are no two sets of events that are equivalent.
This property can be expressed in QPFLTL, and hence
also in QFLTL. The details appear in the appendix.

We use a combinatorial argument to show by contradiction that one cannot express this property
using EFLTL, and also using EPFLTL. The reason is that every prefix of an EFLTL model, extended uniquely with the
auxiliary relations, according to Lemma~\ref{samesame}, can be summarized with a finite number of relations, the ones in
the model, the auxiliary relations and the
assignments satisfying the subformulas. The size of
each such relation is bounde by ${\cal O} (  m ^ N )$
where $m$ is the number of values appearing in the
prefix, and $N$ is the sum of the number of parameters on all the above mentioned relations.
However, the number of different subsets over $m$ values is
$2^m$. This means that with large enough number of different values, EPFLTL can have two prefixes with the same summary. It would then not be able to distinguish between two executions that extend
this prefix, where one of them is extended with a set of values that appeared before and the other with 
a set that already appeared. \qed

@@

%It follows from~(\ref{equiv}) that EFLTL (EPFLTL, %respectively) has the same
%expressive power as QFLTL (EPFLTL, respectively) %restricted to having its
%dynamic 
%quantification to being existential, and prefixing %the
%rest of the formula.
The following follows immediately from Theorem~\ref{theo2} the following:

\begin{corollary}
Constraining the dynamic quantification of QFLTL (QPFLTL) to existential quantification prefixing the
formula strictly weakens its expressive power.
\end{corollary}

@@

\begin{enumerate}
\item Initially, for each subformula $\varphi$,
%$B ( \varphi , \current ) = \bfalse$.
$\current ( \varphi ) := \bfalse$.
\item Observe a new state (as a set of ground predicates) $s_i$ as input. 
\item Let $\old := \current$.
\item Make the following updates for each subformula. If $\varphi$ is
      a subformula of $\psi$ then $\current ( \varphi )$ is updated before 
      $\current ( \psi )$.
\begin{itemize}
  \item $\current ( \true ) := \btrue$.
  \item $\current ( p_k ( a ) )$ := if $R_k [ i ] ( a )$ then
  $\btrue$ else $\bfalse$.
  \item $\current ( p_k ( x ) )$ :=
     $\bddset ( x ,  \{ a \ | \ R_k[i] ( a  ) \} )$.
  \item $\current ( ( \varphi \wedge \psi ) ) := 
  \bddwedge ( \current ( \varphi ) , \current ( \psi ))$.
  \item $\current ( \neg \varphi  ) := \bddneg ( \current ( \varphi ))$.
  \item $\current ( ( \varphi\ \since\ \psi ) ) :=  
  \bddvee ( \current ( \psi  ) , \bddwedge ( \current ( \varphi ) ,
      \old ( ( \varphi \ \since \ \psi ))))$.
%DAP: I removed the \; around \since, since it penetrated the 2nd column!
  \item $\current ( \ominus \; \varphi ) := \old ( \varphi )$.
  \item $\current ( \exists x \; \varphi ) := 
  \bddexists (\langle x_0,\ldots,x_{k-1} \rangle, \current ( \varphi ))$.
\end{itemize}
\item Goto step~2.
\end{enumerate}

>>>>

\item Observe a new event (as a set propositions) $s$ as input. 
\item Let $\old := \current$.
\item Make the following updates for each subformula. If $\varphi$ is
      a subformula of $\psi$ then $\current ( \varphi )$ is updated before 
      $\current ( \psi )$.
\begin{itemize}
  \item $\current ( \true ) := \true$.
  \item $\current (  \varphi \wedge \psi  ) := 
  \current ( \varphi )\  and\ \current ( \psi )$.
  \item $\current ( \neg \varphi  ) := not\ \current ( \varphi )$.
  \item $\current (  \varphi \since \psi  ) :=  
  \current ( \psi  )\ or\ ( \current ( \varphi ) \ and\ 
      \old ( ( \varphi \since  \psi )))$.
%DAP: I removed the \; around \since, since it penetrated the 2nd column!
  \item $\current ( \ominus \; \varphi ) := \old ( \varphi )$.
\end{itemize}
\item If $\current (\eta) = \false$ then
report a violation, otherwise goto step~2.
\end{enumerate}

@@

\noindent
We define\footnote{ Formal semantics 
can also be given by constructing a set of temporal relations extended
with the auxiliary ones
inductively over growing prefixes, as would be done in a detailed proof of Lemma~\ref{samesame}.} the semantics for the EFLTL (EPFLTL)
specification (\ref{EFLTL}) 
by using the following equivalent
QFLTL (QPFLTL, respectively) formula:

@@

\noindent {\bf Sketch of Proof.}
The proof of this theorem includes encoding of a  property that
observes sets of data elements $a$, appearing separately, one per state as $v (a)$,
in between states where $r$ appears. The domain of data elements is unbounded.
%(i.e., $r$ is a relation with
%one parameter: the time/state-index parameter $i$).
The set of $a$-values observed in between two consecutive $r$'s is called a {\em data set}.
The property asserts that there are no two consecutive data sets that are equivalent.
This property can be expressed in QPFLTL. The details appear in the appendix.

We use a combinatorial argument to show by contradiction that one cannot express this property
in EPFLTL. The reason is that every prefix of an a model for an EPFLTL property is extended uniquely with
auxiliary relations, according to Lemma~\ref{samesame}. Each prefix can be summarized by a finite number of relations: the ones in
the model, the auxiliary relations and the
assignments satisfying the subformulas. The size of
each such relation is bounded by ${\cal O} (  m ^ N )$
where $m$ is the number of values appearing in the
prefix, and $N$ is the number of parameters of the relations.
However, the number of different data sets over $m$ values is
$2^m$. This means that with large enough number of different values, each EPFLTL formula over the models of this property can have two prefixes with the same summary where at least one of them has a data set that the other one does not. The semantics of EPFLTL implies that extending
two models with the same summary in the same way would have the same truth value. So, we can extend
the two models with a data set that appears in one of the
prefixes but not  in the other, and the EPFLTL property will not be able to distinguish between them. \qed

@@

%It follows from~(\ref{equiv}) that EFLTL (EPFLTL, %respectively) has the same
%expressive power as QFLTL (EPFLTL, respectively) %restricted to having its
%dynamic 
%quantification to being existential, and prefixing %the
%rest of the formula.

\vspace{0.7ex} From Theorem~\ref{theo2} and Equation~(\ref{equiv}) we immediately obtain:
\begin{corollary}
Restricting the quantification of QPFLTL to existential quantification, strictly weakens its expressive power.
\end{corollary}

@@

\begin{enumerate}
\item Initially, for each subformula $\varphi$ of $\eta$,
%$B ( \varphi , \current ) = \bfalse$.
$\current ( \varphi ) := \bfalse$.
\item Observe a new state (as a set of ground predicates) $s_i$ as input. 
\item Let $\old := \current$.
\item Make the following updates for each subformula. If $\varphi$ is
      a subformula of $\psi$ then $\current ( \varphi )$ is updated before 
      $\current ( \psi )$.
\begin{itemize}
  \item $\current ( \true ) := \btrue$.
  \item $\current ( p_k ( a ) )$ := if $R_k [ i ] ( a )$ then
  $\btrue$ else $\bfalse$.
  \item $\current ( p_k ( x ) )$ :=
     $\bddset ( x ,  \{ a \ | \ R_k[i] ( a  ) \} )$.
  \item $\current ( ( \varphi \wedge \psi ) ) := 
  \bddwedge ( \current ( \varphi ) , \current ( \psi ))$.
  \item $\current ( \neg \varphi  ) := \bddneg ( \current ( \varphi ))$.
  \item $\current ( ( \varphi\ \since\ \psi ) ) :=  
  \bddvee ( \current ( \psi  ) , \bddwedge ( \current ( \varphi ) ,
      \old ( ( \varphi \ \since \ \psi ))))$.
%DAP: I removed the \; around \since, since it penetrated the 2nd column!
  \item $\current ( \ominus \; \varphi ) := \old ( \varphi )$.
  \item $\current ( \exists x \; \varphi ) := 
  \bddexists (\langle x_0,\ldots,x_{k-1} \rangle, \current ( \varphi ))$.
\end{itemize}
\item If $\current (\eta) = \false$ then
report a violation, otherwise goto step~2.
\end{enumerate}
